A generalization of the Bradley-Terry model for draws in chess

### 2019-05-20 inst/collusion_chess_draws.Rnw
work

Bradley-Terry is not directly applicable to Chess for two reasons:
firstly, the white player is at a considerable
advantage~\cite{asdfasdffdsfds}; and secondly, many chess games end in
a draw: white wins about $37.5\%$ of all games, black wins $27.6\%$,
and about $39.4\%$ are drawn.  We consider the first-mover advantage
first, then consider the effect of drawn games.

\paragraph{First-mover advantage} Traditionally White moves first in
any game, and this feature introduces an asymmetry into play.
Historically, we see about 1.35 white wins for every black win.
However, interpretations of this statistic vary and the issue is hotly
debated with opinions ranging from Rowson who compares first-move
advantage to that of serving in tennis, to that of commentators such
as \citeauthor{adorjan} who claims that White's advantage is a
``delusion''.  \cite{asdfasdasfd} goes further, asserting that Black
can exploit the situation by mounting a directed response to White, of
the type seen in analysis of sequential zero-sum games, in which the
first-mover advantage can be positive or negative.

@@

@@

The Method of Support \citep{edwards1972} involves only numerical
optimization (not integration which is computationally expensive).
The analysis presented here uses the framework of the \pkg{hyper2}
package~\citep{hankin2017} as a convenient and unified interface.
Because gradients are available, convergence is generally rapid and
hypothesis-specific constraints on parameter space are, in general,
straightforward to implement.

@@

This likelihood function is easily maximized:

<<>>=
maxp(karpov_kasparov_anand)
@

\begin{figure}[htbp]
\begin{center}
<<ddc,echo=TRUE,fig=TRUE>>=
dotchart(maxp(karpov_kasparov_anand),pch=16,xlim=c(0,0.55))
@
\caption{Maximum likelihood estimate for the strengths of the 3 Grandmasters in Table~\ref{kka} 
\label{maxp_karpov_kasparov_anand} under a likelihood defined as Equation~\ref{likelihood_WD}}
  \end{center}
\end{figure}

and this is shown visually in figure~\ref{maxp_karpov_kasparov_anand},
which suggests a number of interesting and testable statistical
hypotheses.  Firstly, we observe that $D=0$ cannot possibly be the
case: the likelihood for this hypothesis would be identically zero,
owing to the nonzero term in the numerator of the last term of
Equation~\ref{likelihood_WD}.  However, we see that the maximum
likelihood estimate for draw strength $\hat{D}\simeq 0.53$: compare
the na\"ive estimate of $\frac{224}{365}\simeq 0.61$ given by
observing 224 draws among 265 games.  The MLE is lower because the
maximization process automatically adjusts for the different players'
strengths.  Secondly, we may make inferences about $W$, the extra
strength conferred by playing white: it is conceivable that $W=0$ as
this term does not appear on its own in the numerator of any term.  We
may test $H_0\colon W=0$ using the method of support by maximizing the
likelihood subject the constraint $W=0$ and comparing with the
likelihood when the optimization is free.

<<echo=FALSE>>=
small <- 1e-4

@@

<<>>=
ignore3
ignore4
ignore3-ignore4
@

@@

<<>>=
round(maxp(kka_3draws),3)
@

We see considerable variability in proclivity to draw and again this
may be tested with the Method of Support, which gives a $p$-value of
about 0.039, borderline significant: an interesting result, given that
in this case no draw monster appears alone anywhere in the likelihood
function.

<<echo=FALSE,print=FALSE>>=
objective <- function(V){
  karpov <- V[1]
  kasparov <- V[2]
  anand <- V[3]
  white <- V[4]

@@

The method of support allows one to account for proclivity to draw
while accounting for players' strengths and first-move advantage.  We
now consider one famous incident, following the World Chess
Championships 1963, in which Fischer famously and publicly claimed
that the Soviet players colluded.  The accusation was that Soviet
players agreed ahead of time to draw when playing one another, in
order to conserve energy for subsequent matches~\citep{moul2009}.

@@

One plausible explanation for the previ

@@

\noindent (the numerator includes 8 draws and 2 wins, one for Fischer
and one for Korchnoi, both playing white); but the full likelihood
function for the interzonal has 297 terms, including $D^{110}$.  To
investigate the possibility of Soviet collusion we introduce a new
draw entity: a soviet Draw monster $D_s$, who wins when two Soviet
players draw, and is inactive whenever a game with at least one
non-Soviet player is drawn.  Then $H_N\colon D_s=D$ corresponds to no
collusion and the likelihood function for table~\ref{interzonal} would
be

@@

The maximum likelihood estimate is $D\simeq 0.043, D_s=0.354$; a
constrained maximization gives a likelihood ratio of about 5096, or a
$p$-value of about $3.6\times 10^{-5}$.  Indeed we could go further
and plot a profile likelihood curve for the log-odds ratio
$\ln\left(D_s/D\right)$ which, if Dirichlet, would be accurately
parabolic over a wide range (OHAGAN AND FORSTER, p343).  This is
illustrated in Figure~\ref{DOESNOTEXIST}, that shows a credible
interval of about HAVENOTDONETHISET.

@@

\begin{figure}[htbp]
\begin{center}
<<proflike,echo=TRUE,fig=TRUE>>=
plot(x,jj-max(jj),type='b',xlab="log-odds",ylab="support")
abline(h=c(0,-2))
@
\caption{Profile likelihood \label{proflike} for odds ratio
  $\log\left(D_s/D\right)$: note the accurately parabolic figure in
  the region plotted}
  \end{center}
\end{figure}

@@

In this context we note, following \cite{langer2013}, that a player
may well adopt a riskier strategy after a sequence of defeats,
adopting a ``nothing-to-lose'' mentality.  Such a phenomenon would be
detectable, at least in principle, by generalizing the personalised
draw likelihood function \ref{KKA_personalised_drawmonster} and
seeking evidence for the draw monster's strength changing as a
tournament progresses: it would become stronger for a player on a
winning streak, and weaker for a losing player.

>>>>

Bradley-Terry is not directly applicable to Chess for two reasons:
firstly, the white player appears to be at a considerable advantage;
and secondly, many chess games end in a draw: white wins about
$37.5\%$ of all games, black wins $27.6\%$, and about $39.4\%$ are
drawn.  We consider the first-mover advantage first, then consider the
effect of drawn games.

\paragraph{First-mover advantage} Traditionally White moves first in
any game, and this feature introduces an asymmetry into play.
Historically, we see about 1.35 white wins for every black win.
However, interpretations of this statistic vary and the issue is hotly
debated with opinions ranging from that of~\cite{adams2007}---``white
to play and win''---to~\citet{adorjan1988}---``black is OK!''.

@@

\subsection{Previous work}

\cite{davidson1970} uses a threshhold approach to interpret a draw as
performance of the players the same to within a threshhold.  Using a
logistic expression for the difference in players' performance, he
effectively uses, in current notation, $D=\nu\sqrt{p_ip_j}$.  Here we
note that such a model automatically give stronger players a higher
draw potential, and also note that it is difficult to account for
first-player advantage in this framework.  Such an approach is
numerically challenging in the current framework as the unit-sum
constraint becomes nonlinear and likelihood maximization becomes
difficult.

@@

The Method of Support \citep{edwards1972} involves only numerical
optimization (not integration which is computationally expensive).
The analysis presented here uses the framework of the \pkg{hyper2}
package~\citep{hankin2017} as a convenient and unified interface.
Because gradients are available, convergence to the evaluate is
generally rapid and hypothesis-specific constraints on parameter space
are, in general, straightforward to implement.

@@

<<echo=FALSE,print=FALSE,results=hide>>=
maxp(karpov_kasparov_anand)
@

\begin{figure}[t]
\begin{center}
<<ddc,echo=FALSE,fig=TRUE>>=
dotchart(maxp(karpov_kasparov_anand),pch=16,xlim=c(0,0.55))
@
\caption{Maximum likelihood estimate for the strengths of the 3 Grandmasters in Table~\ref{kka} 
\label{maxp_karpov_kasparov_anand} under a likelihood defined as Equation~\ref{likelihood_WD}}
  \end{center}
\end{figure}

This likelihood function is easily maximized numerically
and this is shown visually in figure~\ref{maxp_karpov_kasparov_anand},
which suggests a number of interesting and testable statistical
hypotheses.  Firstly, we observe that $D=0$ cannot possibly be the
case: the likelihood for this hypothesis would be identically zero,
owing to the nonzero term in the numerator of the last term of
Equation~\ref{likelihood_WD}.  However, we see that the maximum
likelihood estimate for draw strength $\hat{D}\simeq 0.53$: compare
the na\"ive estimate of $\frac{224}{365}\simeq 0.61$ given by
observing 224 draws among 265 games.  The MLE is lower because the
maximization process automatically adjusts for the different players'
strengths.  Secondly, we may make inferences about $W$, the extra
strength conferred by playing white: it is conceivable that $W=0$ as
this term does not appear on its own in the numerator of any term.  We
may test $H_0\colon W=0$ using the method of support by maximizing the
likelihood subject the constraint $W=0$ and comparing with the
likelihood when the optimization is free.

<<maxlikekka,echo=FALSE,print=FALSE,results=hide>>=
small <- 1e-4

@@

<<ignore34diff,echo=FALSE,print=FALSE,results=hide>>=
ignore3
ignore4
ignore3-ignore4
@

@@

\begin{figure}[t]
\begin{center}
<<echo=FALSE,fig=TRUE>>=
dotchart(maxp(kka_3draws),pch=16)
@ 
\caption{Maximum likelihood estimate for \label{kka_3draws} equation \ref{KKA_personalised_drawmonster} equation} 
  \end{center}
\end{figure}

In figure~\ref{kka_3draws}, we see considerable variability in
proclivity to draw and again this may be tested with the Method of
Support, which gives a $p$-value of about 0.039, borderline
significant: an interesting result, given that in this case no draw
monster appears alone anywhere in the likelihood function.

<<echo=FALSE,print=FALSE,results=hide>>=
objective <- function(V){
  karpov <- V[1]
  kasparov <- V[2]
  anand <- V[3]
  white <- V[4]

@@

The method of support allows one to account for proclivity to draw
while accounting for players' strengths and first-move advantage.  We
now consider one famous incident, following the World Chess
Championships 1963, in which Fischer famously and publicly claimed
that the Soviet players colluded.  The accusation was that Soviet
players agreed ahead of time to draw when playing one another, in
order to conserve energy for subsequent matches.

\citet{moul2009} consider this accusation from a statistical
perspective and present a detailed analysis of Cura\c{c}ao.  Unlike
the present analysis, which uses a dataset restricted to game result
(win/loss/draw), \citet{moul2009} included number of moves in each
game\footnote{The reasoning was that hard-fought games that happened
  to end in a draw are more likely to last longer than collusive or
  pre-arranged draws.} in their analysis, and used a dataset of 7355
games played in the period 1940-1978 to calibrate the ability of the
players.  In contrast, the analysis presented uses a dataset
restricted to game outcome (win-loss-draw) in the competition itself.

@@

One plausible and charitable explanation for this result is that
Soviet players have a higher propensity to draw than western players.
We can investigate this by considering the interzonal tournament which
took place in Stockholm in 1962.

@@

\noindent (the numerator includes 8 draws and 2 wins, one for Fischer
and one for Korchnoi, both playing white); but the full likelihood
function for the interzonal has 297 terms, including $D^{110}$.  To
investigate the possibility of Soviet collusion we introduce a new
draw entity: a soviet Draw monster $D_s$, who wins when two Soviet
players draw, and is inactive whenever a game has at least one
non-Soviet player.  Then $H_N\colon D_s=D$ corresponds to no collusion
and the likelihood function for table~\ref{interzonal} would be

@@

The maximum likelihood estimate is $D\simeq 0.043, D_s=0.354$; a
constrained maximization gives a likelihood ratio of about 5096, or a
$p$-value of about $3.6\times 10^{-5}$.  Indeed we could go further
and plot a profile likelihood curve for the log-odds ratio
$\ln\left(D_s/D\right)$ which, if Dirichlet, would be accurately
parabolic over a wide range~\cite[page 343]{ohagan2004}.  This is
illustrated in Figure~\ref{proflike}, that shows a credible interval
of about 1.8-2.4.

@@

\begin{figure}[htbp]
\begin{center}
<<proflike,echo=FALSE,fig=TRUE>>=
plot(x,jj-max(jj),type='b',xlab="log-odds",ylab="support")
abline(h=c(0,-2))
@
\caption{Profile likelihood \label{proflike} for odds ratio
  $\log\left(D_s/D\right)$: note the accurately parabolic figure in
  the region plotted}
  \end{center}
\end{figure}

@@

In this context we note, following \cite{langer2013}, that a player
may well adopt a riskier strategy after a sequence of defeats,
adopting a ``nothing-to-lose'' mentality.  Such a phenomenon would be
detectable, at least in principle, by generalizing the personalised
draw likelihood function \ref{KKA_personalised_drawmonster} and
seeking evidence for the draw strength changing as a tournament
progresses: it would become stronger for a player on a winning streak,
and weaker for a losing player.
